---
title: "Preparing data and choosing sampler settings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing data and choosing sampler settings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article focuses on the decisions that happen before calling
`pgmm_rjmcmc()`: matrix orientation, scaling, missing values, latent-factor
dimension, cluster range, and starting covariance model. The goal is to make
applied use less ambiguous.

## Matrix orientation

`bpgmm` expects a numeric matrix with variables in rows and observations in
columns. Many R data sets use the opposite convention, so transpose after
selecting numeric variables.

```{r}
library(bpgmm)

iris_numeric <- as.matrix(iris[, 1:4])
iris_labels <- as.integer(iris$Species)

dim(iris_numeric)

X <- t(iris_numeric)
dim(X)
```

Rows now correspond to variables and columns correspond to observations.

```{r}
rownames(X)
X[, 1:3]
```

## Check finite numeric inputs

The sampler requires finite numeric values. Handle missing values before
fitting. Common choices include complete-case filtering, domain-specific
imputation, or fitting the model to a subset of variables with reliable
measurements.

```{r}
all(is.finite(X))
storage.mode(X)
```

## Scale variables

Mixture models are sensitive to measurement scale. If variables are measured in
different units, standardizing each variable is usually a sensible default.

```{r}
X_scaled <- t(scale(t(X)))

round(rowMeans(X_scaled), 6)
round(apply(X_scaled, 1, sd), 6)
```

The package does not scale internally because some scientific applications need
the original measurement scale. Scaling should be an explicit analysis choice.

## Choose `q_new`

`q_new` is the latent-factor dimension assigned to newly proposed clusters. It
controls the dimension of the factor-analyzer part of the covariance model.

Useful starting points:

- `q_new = 1` for very small examples or when covariance structure should be
  simple.
- `q_new = 2` or `3` for moderate-dimensional data.
- Larger values only when the number of variables and observations support the
  extra covariance flexibility.

The value should be smaller than the number of observed variables.

```{r}
p <- nrow(X_scaled)
q_new <- min(2, p - 1)
q_new
```

## Choose `m_range`

`m_range` is the allowed cluster-number range. A wide range gives RJMCMC more
freedom, but also increases the model space. Start with a scientifically
reasonable range, then assess sensitivity.

```{r}
m_init <- 3
m_range <- c(1, 5)
```

For data with a known reference label, such as `iris`, this range is easy to
check. In real clustering problems, use domain knowledge and exploratory plots.

```{r, fig.width = 5.5, fig.height = 4.2, fig.alt = "Scatter plot of scaled iris variables colored by species."}
species_cols <- c("#0072B2", "#D55E00", "#009E73")
plot(
  X_scaled[1, ], X_scaled[2, ],
  col = species_cols[iris_labels],
  pch = 19,
  xlab = rownames(X_scaled)[1],
  ylab = rownames(X_scaled)[2],
  main = "Scaled iris data",
  asp = 1
)
legend(
  "topleft",
  legend = levels(iris$Species),
  col = species_cols,
  pch = 19,
  bty = "n"
)
```

## Choose the starting covariance model

The three-letter model labels describe whether loading matrices and noise
covariances are shared across clusters. `UUU` is flexible; `CCC` is more
constrained. A flexible starting model is often reasonable when using
`v_step = 1`, because the sampler can move across covariance structures.

```{r}
model_to_constraint("UUU")
constraint_to_model(c(1, 1, 1))
```

## A prepared call

The call below is intentionally not evaluated in the vignette because a real
analysis should use longer chains and repeated runs. It shows how the prepared
objects fit into the package interface.

```{r, eval = FALSE}
fit <- pgmm_rjmcmc(
  X = X_scaled,
  m_init = m_init,
  m_range = m_range,
  q_new = q_new,
  burn = 1000,
  niter = 5000,
  constraint = "UUU",
  m_step = 1,
  v_step = 1,
  split_combine = 1,
  verbose = FALSE
)

summarize_pgmm_rjmcmc(fit, true_cluster = iris_labels)
```

## Practical checklist

Before fitting:

- verify that `X` is numeric with variables in rows;
- remove or impute missing and non-finite values;
- decide whether variables should be scaled;
- choose a scientifically plausible `m_range`;
- choose `q_new` smaller than the number of variables;
- use multiple chains for applied analyses.
