---
title: "Worked examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article shows a complete, runnable `bpgmm` workflow. The example is small
enough to build quickly on CRAN and pkgdown, but it uses the same functions as a
larger analysis.

## Simulate a small clustering problem

`pgmm_rjmcmc()` expects a numeric matrix with variables in rows and observations
in columns. The code below creates two compact clusters in two dimensions.

```{r}
library(bpgmm)

set.seed(2026)

X <- cbind(
  matrix(rnorm(8, mean = -2, sd = 0.2), nrow = 2),
  matrix(rnorm(8, mean = 2, sd = 0.2), nrow = 2)
)
known_labels <- rep(1:2, each = 4)

dim(X)
known_labels
```

## Choose a covariance model

The paper describes eight covariance structures using three-letter PGMM labels:
`CCC`, `CCU`, `CUC`, `CUU`, `UCC`, `UCU`, `UUC`, and `UUU`. Use
`model_to_constraint()` when calling the sampler, and `constraint_to_model()`
when reading stored results.

```{r}
constraint <- model_to_constraint("UUU")
constraint

constraint_to_model(constraint)
```

## Fit a short RJMCMC chain

This vignette uses only three posterior iterations so it runs quickly. For
research use, increase `burn` and `niter`, and set `m_step = 1` and `v_step = 1`
when you want RJMCMC model selection for the number of clusters and covariance
structure.

```{r}
fit_log <- capture.output({
  fit <- pgmm_rjmcmc(
    X = X,
    m_init = 2,
    m_range = c(1, 3),
    q_new = 1,
    burn = 1,
    niter = 3,
    constraint = constraint,
    m_step = 0,
    v_step = 0,
    verbose = FALSE
  )
})
tail(fit_log, 1)
```

The returned object stores posterior samples for allocations, covariance
constraints, component parameters, and hyperparameters.

```{r}
names(fit)
length(fit$allocation_samples)
```

## Summarize posterior samples

`summarize_pgmm_rjmcmc()` returns the modal allocation, posterior counts for the
number of clusters, posterior counts for covariance models, and optionally an
adjusted Rand index.

```{r}
fit_summary <- summarize_pgmm_rjmcmc(fit, true_cluster = known_labels)

fit_summary$allocation
fit_summary$n_clusters
fit_summary$n_constraints
fit_summary$ari
```

## Scale the example to real data

For a real analysis, keep the same workflow but use larger sampler settings.
The exact values depend on data size and convergence diagnostics.

```{r, eval = FALSE}
fit <- pgmm_rjmcmc(
  X = your_data_matrix,
  m_init = 2,
  m_range = c(1, 10),
  q_new = 4,
  burn = 5000,
  niter = 15000,
  constraint = model_to_constraint("UUU"),
  m_step = 1,
  v_step = 1,
  split_combine = 1,
  verbose = FALSE
)

summarize_pgmm_rjmcmc(fit, true_cluster = known_labels)
```
