---
title: "Getting started with bpgmm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bpgmm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`bpgmm` fits Bayesian parsimonious Gaussian mixture models for model-based
clustering. It targets three posterior inference goals described by Lu, Li, and
Love (2021): the partition of observations, the number of clusters, and the
cluster covariance structure.

## Fit a model

Input data should be a numeric matrix with variables in rows and observations
in columns.

```{r}
library(bpgmm)

fit <- pgmm_rjmcmc(
  X = X,
  mInit = 2,
  mVec = c(1, 6),
  qnew = 2,
  burn = 100,
  niter = 1000,
  Mstep = 1,
  Vstep = 1
)
```

Use `Mstep = 1` to allow RJMCMC updates for the number of clusters. Use
`Vstep = 1` to allow RJMCMC updates across the covariance-constraint models.

## Summarize posterior samples

```{r}
summary <- summarize_pgmm_rjmcmc(fit)

summary$Zalloc
summary$nCluster
summary$nConstraint
```

If a reference partition is available, pass it as `trueCluster` to calculate
the adjusted Rand index.

```{r}
summary <- summarize_pgmm_rjmcmc(fit, trueCluster = known_labels)
summary$ari
```

## Work with covariance constraints

The paper uses the eight PGMM covariance model labels `CCC`, `CCU`, `CUC`,
`CUU`, `UCC`, `UCU`, `UUC`, and `UUU`. The package stores these internally as
three-number constraint vectors.

```{r}
model_to_constraint("UUU")
constraint_to_model(c(1, 0, 0))
```

## Deprecated names

The old names `pgmmRJMCMC()`, `summarizePgmmRJMCMC()`, and
`summerizePgmmRJMCMC()` remain available as compatibility wrappers, but new code
should use `pgmm_rjmcmc()` and `summarize_pgmm_rjmcmc()`.

## Citation

If you use `bpgmm` in published work, please cite the package and the
methodology paper:

```{r}
citation("bpgmm")
```

Lu, X., Li, Y., & Love, T. (2021). On Bayesian Analysis of Parsimonious
Gaussian Mixture Models. *Journal of Classification*, 38, 576-593.
https://doi.org/10.1007/s00357-021-09391-8

