---
title: "Getting started with bpgmm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bpgmm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`bpgmm` fits Bayesian parsimonious Gaussian mixture models for model-based
clustering. It targets three posterior inference goals described by Lu, Li, and
Love (2021): the partition of observations, the number of clusters, and the
cluster covariance structure.

For the model equations, prior distributions, PGMM covariance labels, and
RJMCMC move types, see `vignette("model-and-sampler", package = "bpgmm")`.

## Fit a model

Input data should be a numeric matrix with variables in rows and observations
in columns. The small example below has two variables and eight observations.
It is intentionally short so the vignette runs quickly; use larger `burn` and
`niter` values for real analysis.

```{r}
library(bpgmm)

set.seed(2026)

X <- cbind(
  matrix(rnorm(8, mean = -2, sd = 0.2), nrow = 2),
  matrix(rnorm(8, mean = 2, sd = 0.2), nrow = 2)
)
known_labels <- rep(1:2, each = 4)

fit_log <- capture.output({
  fit <- pgmm_rjmcmc(
    X = X,
    mInit = 2,
    mVec = c(1, 3),
    qnew = 1,
    burn = 1,
    niter = 3,
    constraint = model_to_constraint("UUU"),
    Mstep = 0,
    Vstep = 0,
    verbose = FALSE
  )
})
tail(fit_log, 1)
```

The arguments control the sampler:

- `mInit` is the initial number of clusters.
- `mVec` is the allowed range for the number of clusters.
- `qnew` is the number of latent factors for a new cluster.
- `constraint` selects the starting covariance model.
- `Mstep = 1` allows RJMCMC updates for the number of clusters.
- `Vstep = 1` allows RJMCMC updates across covariance-constraint models.

A longer run for applied work might look like this:

```{r, eval = FALSE}
fit <- pgmm_rjmcmc(
  X = methylation_matrix,
  mInit = 2,
  mVec = c(1, 6),
  qnew = 2,
  burn = 100,
  niter = 1000,
  Mstep = 1,
  Vstep = 1,
  verbose = FALSE
)
```

## Summarize posterior samples

```{r}
summary <- summarize_pgmm_rjmcmc(fit, trueCluster = known_labels)

summary$Zalloc
summary$nCluster
summary$nConstraint
summary$ari
```

If a reference partition is available, pass it as `trueCluster` to calculate
the adjusted Rand index.

## Work with covariance constraints

The paper uses the eight PGMM covariance model labels `CCC`, `CCU`, `CUC`,
`CUU`, `UCC`, `UCU`, `UUC`, and `UUU`. The package stores these internally as
three-number constraint vectors.

```{r}
model_to_constraint("UUU")
constraint_to_model(c(1, 0, 0))
```

## Deprecated names

The old names `pgmmRJMCMC()`, `summarizePgmmRJMCMC()`, and
`summerizePgmmRJMCMC()` remain available as compatibility wrappers, but new code
should use `pgmm_rjmcmc()` and `summarize_pgmm_rjmcmc()`.

## Citation

If you use `bpgmm` in published work, please cite the package and the
methodology paper:

```{r}
citation("bpgmm")
```

Lu, X., Li, Y., & Love, T. (2021). On Bayesian Analysis of Parsimonious
Gaussian Mixture Models. *Journal of Classification*, 38, 576-593.
https://doi.org/10.1007/s00357-021-09391-8
