---
title: "Model and sampler details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model and sampler details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article connects the notation in Lu, Li, and Love (2021) to the
`bpgmm` interface. The goal is to make clear what the package is fitting, what
the covariance labels mean, and how the RJMCMC switches change the sampler.

## Observation model

Let \(X = (x_1, \ldots, x_n)\) be a matrix of continuous observations with
\(p\) variables and \(n\) observations. `bpgmm` expects this same orientation:
variables in rows and observations in columns.

For cluster \(k\), the mixture of factor analyzers model writes

\[
x_i = \mu_k + \Lambda_k y_{ki} + \epsilon_{ki},
\]

where

\[
y_{ki} \sim N(0, I_{q_k}), \qquad
\epsilon_{ki} \sim N(0, \Psi_k).
\]

After integrating out the latent factor \(y_{ki}\), the cluster-specific
covariance is

\[
\Sigma_k = \Lambda_k \Lambda_k^\top + \Psi_k.
\]

The mixture density is

\[
f(x_i) = \sum_{k = 1}^m \tau_k
  N(x_i \mid \mu_k, \Lambda_k \Lambda_k^\top + \Psi_k),
\]

where \(\tau_k\) is the mixture weight and \(m\) is the number of clusters.

In `pgmm_rjmcmc()`:

- `X` is the \(p \times n\) data matrix.
- `m_init` is the starting value of \(m\).
- `m_range` is the allowed range of \(m\).
- `q_new` is the factor dimension assigned to a newly proposed cluster.
- `constraint` is the starting covariance model.

## PGMM covariance constraints

The paper uses three letters to describe the covariance structure. Each letter
is either `C` for constrained or `U` for unconstrained.

| Letter | Meaning when `C` | Meaning when `U` |
|---|---|---|
| 1 | all clusters share one loading matrix \(\Lambda\) | each cluster has \(\Lambda_k\) |
| 2 | all clusters share one noise covariance \(\Psi\) | each cluster has \(\Psi_k\) |
| 3 | noise covariance is isotropic, \(\Psi_k = \psi_k I_p\) | noise covariance is diagonal |

The eight PGMM models are:

```{r}
library(bpgmm)

models <- c("CCC", "CCU", "CUC", "CUU", "UCC", "UCU", "UUC", "UUU")
data.frame(
  model = models,
  constraint = vapply(
    models,
    function(x) paste(model_to_constraint(x), collapse = ","),
    character(1)
  )
)
```

The package uses the numeric encoding internally: `1` means constrained and
`0` means unconstrained.

```{r}
model_to_constraint("UUU")
constraint_to_model(c(1, 1, 0))
```

## Priors and posterior updates

The supplement gives the natural conjugate priors used by the MCMC updates.
In compact form, the main priors are:

\[
\tau \sim \mathrm{Dirichlet}(\gamma, \ldots, \gamma),
\]

\[
y_{ki} \sim N(0, I_{q_k}),
\]

\[
\mu_k \sim N(\bar{x}, \alpha_1^{-1} \Psi_k),
\]

\[
\lambda_{kj} \sim N(0, \alpha_2^{-1} \Psi_k),
\]

and

\[
\psi_{kj} \sim \mathrm{IG}(\delta, \beta).
\]

The package samples the hyperparameters \(\alpha_1\), \(\alpha_2\), and
\(\beta\), with gamma hyperpriors controlled by:

- `d_vec`: shape parameters.
- `s_vec`: rate parameters.
- `delta`: inverse-gamma shape for the noise covariance.
- `ggamma`: Dirichlet concentration for mixture weights.

The allocation update uses the posterior probability

\[
p_{ki} =
\frac{
  \tau_k N(x_i \mid \mu_k, \Psi_k + \Lambda_k\Lambda_k^\top)
}{
  \sum_{\ell = 1}^{m}
  \tau_\ell N(x_i \mid \mu_\ell, \Psi_\ell + \Lambda_\ell\Lambda_\ell^\top)
}.
\]

Then

\[
(z_{i1}, \ldots, z_{im}) \mid X, \Theta, H
  \sim \mathrm{Multinomial}(1, p_{1i}, \ldots, p_{mi}).
\]

This is implemented in the package's allocation update and is also why the
joint posterior includes the product of allocated mixture weights:

\[
p(Z \mid \tau) = \prod_{i = 1}^{n} \tau_{z_i}.
\]

## RJMCMC moves

RJMCMC is used because the number of parameters changes when the number of
clusters or the covariance constraint changes.

The cluster-number moves are:

| Move | Purpose |
|---|---|
| stay | update parameters without changing \(m\) |
| birth | add a new empty cluster |
| death | remove an empty cluster |
| split | split one occupied cluster into two clusters |
| combine | merge two occupied clusters |

The covariance-structure moves toggle one constraint at a time:

- toggle whether \(\Lambda_k\) is shared across clusters;
- toggle whether \(\Psi_k\) is shared across clusters;
- toggle whether \(\Psi_k\) is isotropic or diagonal.

In `pgmm_rjmcmc()`:

- `m_step = 1` enables birth/death moves for \(m\).
- `split_combine = 1` adds split/combine moves when `m_step = 1`.
- `v_step = 1` enables covariance-constraint moves.

## Minimal runnable example

The following example is deliberately small so the vignette can run quickly.
For real analyses, increase `burn` and `niter`.

```{r}
set.seed(2026)

X <- cbind(
  matrix(rnorm(8, mean = -2, sd = 0.2), nrow = 2),
  matrix(rnorm(8, mean = 2, sd = 0.2), nrow = 2)
)
known_labels <- rep(1:2, each = 4)

fit <- pgmm_rjmcmc(
  X = X,
  m_init = 2,
  m_range = c(1, 3),
  q_new = 1,
  burn = 1,
  niter = 3,
  constraint = model_to_constraint("UUU"),
  m_step = 0,
  v_step = 0,
  verbose = FALSE
)

fit_summary <- summarize_pgmm_rjmcmc(fit, true_cluster = known_labels)
fit_summary$n_clusters
fit_summary$n_constraints
fit_summary$ari
```

The examples above keep `m_step = 0` and `v_step = 0` to make the vignette fast.
A full model-selection run uses the same interface:

```{r, eval = FALSE}
fit <- pgmm_rjmcmc(
  X = your_data_matrix,
  m_init = 5,
  m_range = c(1, 10),
  q_new = 4,
  burn = 5000,
  niter = 15000,
  constraint = model_to_constraint("UUU"),
  m_step = 1,
  v_step = 1,
  split_combine = 1,
  verbose = FALSE
)
```

## Reading posterior output

The sampler returns posterior samples as lists. The most commonly used fields
are:

| Field | Meaning |
|---|---|
| `allocation_samples` | sampled cluster allocations \(Z\) |
| `constraint_samples` | sampled PGMM covariance constraints |
| `tau_samples` | sampled mixture weights \(\tau\) |
| `mean_samples` | sampled cluster means \(\mu_k\) |
| `lambda_samples` | sampled loading matrices \(\Lambda_k\) |
| `psi_samples` | sampled noise covariances \(\Psi_k\) |

Use `summarize_pgmm_rjmcmc()` for posterior summaries:

```{r}
names(fit)
names(fit_summary)
```

## Citation

If you use these methods, cite:

Lu, X., Li, Y., & Love, T. (2021). On Bayesian Analysis of Parsimonious
Gaussian Mixture Models. *Journal of Classification*, 38, 576-593.
https://doi.org/10.1007/s00357-021-09391-8

In R:

```{r}
citation("bpgmm")
```
